\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization \thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602 (phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}

### 2010-03-30 honors_thesis/journal_paper/paper.tex
kevin finished second pass

Particle swarm optimization (PSO) has previously been parallelized only by 
adding more particles to the swarm or by parallelizing the evaluation of the
objective function.  However, in many cases this approach does not scale to 
very large populations sizes (on very large clusters).
In response to this issue, we take inspiration from 
speculative execution commonly performed in modern processors and propose
Speculative Evaluation in PSO (SEPSO).

In SEPSO future positions of the particles are
speculated and evaluated in parallel with the evaluation of current positions.
We show that the set of speculative evaluations will always include the position and corresponding
evaluation that would normally have occurred at the next iteration.
Thus SEPSO performs two
iterations of PSO at once, allowing it to achieve better fitness faster.
Not only can SEPSO can achieve better fitness in the same number of iterations,
it also reaches better fitness for the same number of total function evaluations
in large scale parallel environments.

While Speculative Evaluation in PSO is an intriguing idea, its
usefulness is limited unless some slight modifications are made to the PSO
algorithm.  By relaxing the requirement that speculative evaluation exactly
reproduce the behavior of the original PSO, we can make better use of the information
obtained by the speculative evaluations.
Using these algorithmic modifications, we see dramatic improvements in the
effectiveness of the algorithm. 
\end{abstract}

@@

These previous parallel techniques use additional processors to increase the
number of particles in the swarm, either by adding individual particles or by
adding entire new sub-swarms but in all cases
the number of processors never exceeds the number of particles.
The number of iterations of PSO that the
algorithms can perform is thus inherently limited by the time it takes to
evaluate the objective function---additional processors add more particles, not
more iterations.

The purpose of this paper is to consider PSO parallelization strategies for
clusters of thousands of processors and functions for which a single
evaluation will take at least minutes, but probably hours or perhaps days.
Our purpose is to explore the question ``Since PSO is often run with
a swarm of size of 50-100, what would one do if a few thousand processors were available?''

For many functions there comes a point of diminishing returns with respect to adding particles.
In \fig{basic-sphere} we show the value obtained after 20,000 function evaluations (not iterations)
as a function of swarm size
for the well-know benchmark function Sphere (20 dimensions, reporting the average of thirty runs). 
Increasing the swarm size from 10 to 50 has a significant effect value obtained
however, increasing the swarm size from 100 to 200 makes the algorithm less efficient,
reducing the progress the algorithm makes per-evaluation.
For this reason, research recommends the use a swarm size of 50 for this function
~\cite{bratton-sis07}.
Thus, in at least some cases, added particles indefinitely will not yield an efficient implementation.

@@

To solve this problem, we take inspiration from speculative execution
techniques commonly used in processors.  Modern processors, when faced with a
branch on which they must wait (e.g. a memory cache miss), guess which way the
branch will go and start executing.  If the processor guesses right, execution
is much farther ahead than if it had just idly waited on the branch.  If it
guesses wrong, execution restarts where it would have been anyway.
This is called speculative execution or branch prediction.

The idea of speculative execution can be applied to some types of optimization
algorithms.  When the next sampling position of the algorithm does not depend
on the value of the objective function at the current position,
future positions can be speculated and evaluated in parallel with the current
position.
This is trivially true for optimization by random sampling, for example.
The next random sample does not depend on the previous point nor on the value
of the objective function at that point in any way.
It depends only on a random number.
There is no need to wait to see
what objective function value was obtained at the previous point,
evaluation of the object function at the next point can begin as soon as the next random number is drawn.
In other optimization algorithms, the next position depends on the previous position, but not the
value obtained there, a random walk for example.
In this case
the next position to be evaluated
depends only on the previous position and a random number.
In this case, as it was in the case of random sampling,
the object function evaluation of the next position can be done in parallel
because it 
does not depend on the \emph{value} of the objective function at the previous position.
There are an infinite number of possible next positions, but only one given the draw of the random number.

In still other cases, the next position to be evaluated may depend on the current position and some
discrete valued (binary, for example) test on the value obtained at the current position.
For example, consider a greedy random walk, in which the state only advances if a better
position is found.
That is, if the current value is better than the previous value,
use the sum of the current position and the random number as the next position.
Otherwise, computer the next position using the sum of the previous position and the random number.
In this case the objective function can be evaluated at three positions:
(1) the value of the objective function at the current position,
(2) the value at the position if the test turns out to be true and (3) the value at the position
if the test turns out to be false; all in parallel.
In this example it is important to note that we do not need the value of the objectve function at the current
point to to reduce the infinite number of possible next postions to just two alternatives.
Given the random sample, the next point detments only on the binary test and the
random sample is, by definition, idependent of the positions, and the vaues obtained.
The choice of the next position can be made post-hoc, after the value at the current position is obtained.
For all but the most computationally-trivial objective functions, this approach saves almost half of the
``wall-clock'' time.
This approach has been applied to simulated annealing~\cite{witte-tpds91} which like the greedy random walk,
depends on positions and a binary test.

This approach can not be used in gradient methods where the
step size for sampling typically depends on the gradient and therefor the position and 
a set of infinitly many posible \emph{values} outcomes
obtained at the current position (not binary or even finite).

@@

In this paper we show that the results of standard PSO can be reproduced \emph{exactly}, two iterations at a time, using
a speculative evaluation approach like the ones described above. We prove that this approach
produces exactly equivalent results, and demonstrate this by implementation.
The resulting implementation runs efficiently on large clusters where the number of processors is much larger than
a typical or reasonable number of particles. Producing better results in less ``wall-clock'' time and,
surprisingly, with fewer total function evaluations.

Furthermore, we show that if we relax the requirements of the algorithm, no
longer demanding that it strictly reproduce the exact behavior of standard PSO, we
can introduce new speculative techniques that often out-perform PSO.
These variants make better use of the extra information obtained from
the extra exploration made by the speculative function evaluations.
We also explore the idea that, like branch prediction in processors, we
need not speculatively evaluate \emph{all} possible future positions,
we can accelerate the algorithm even if we are just \emph{likely} to have guessed right.
By pruning the speculation to just paths that are statistically likely to reproduce the paths that are equivelent to PSO we can also 
speculating several iterations ahead instead of just one.
We also consider several recovery strategies for cases where
the pruned set of speculative evaluations does not contain the evaluation that standard PSO would have done.

@@

At each iteration ($t+1$) of the algorithm, the position $\ppos_t$ and velocity
$\pvel_t$ of each particle are updated as follows:
\begin{align}
\nonumber
	\pvel_{t+1} &=
		\constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\label{eq:velupdate}
			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
		\bigr] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1}
\end{align}
where \( \nURand_{t} \) and \( \nURand_{t} \) are vectors of independent
and identically distributed random numbers drawn from a uniform
distribution, the \( \otimes \) operator is an element-wise vector
multiplication, $\pbest$ (called personal best) is the best position the
current particle has seen, and $\nbest$ (called neighborhood best) is the best
position the neighbors of the current particle have seen~\cite{bratton-sis07}.  \( \ncoeff \), \(
\pcoeff \), and \( \constriction \) are parameters with prescribed values
required to ensure convergence (2.05, 2.05, and .73,
respectively)~\cite{clerc-tec02}. Note also that we use a subscript for the 
time step $t$ or $t+1$ and have moved all other qualifiers,
such as the personal qualifier ``$\pers$'' to the superscript position.
Superscripts can be miss read as exponents, but they are qualifiers throughout this paper.

There are many ways of defining the neighbors of a particle. 
Changing the
neighborhood topology has a significant effect on the performance of the algorithm and 
this effect varies widely by function type, as would be expected by the No Free Lunch
theorem~\cite{wolpert-tec97}.
The two most common topologies used in the
literature are the Ring topology and the Complete
topology~\cite{bratton-sis07}.  In the Ring topology, each particle has one
or more neighbors to either side of it; in the Complete topology, every particle is a
neighbor to every other particle.  In all topologies, a particle is also a
neighbor to itself in that its own position is considered when updating the
particle's personal best, $\nbest$.  Thus with $p$ particles, using the Ring topology each
particle has three neighbors: $i-1$, $i$ (itself), and $i+1$.  With the Complete
topology, each particle has $p$ neighbors.
% TOD USE THIS LATER!
% In this paper we use these
% topologies, as well as a parallel adaptation of the Complete topology, called
% Random, that has been shown to simulate the behavior of Complete with far less
% communication~\cite{mcnabb-cec09}.  In the Random topology, each particle
% randomly picks two other particles to share information with at each iteration,
% along with itself.  Thus in both the Ring and the Random topologies, all
% particles have 3 neighbors.

@@

To address this issue we present here a very different use of additional processors.
We take inspiration from speculative execution in processors and instead of adding particles
we will employ a speculative evaluation approach like those described in the introduction. 
This allows us to perform iterations two at a time.
To do so we must first re-factor PSO such that the determination of the $\pbest$ and $\nbest$,
which depend upon the value of the objective function, is sepearate from the rest of the computation.

%First consider formula for the \emph{next} iteration of PSO, which is identical
%to Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate} except the time
%subscripts are incremented.  As noted above and
For simplicity in this
explanation, we restrict our discussion here to the case of a Ring topology
with two neighbors which we will call the ``right neighbor'' and ``left
neighbor'' neighbors identified for each particle as $\rightn$ and $\leftn$
respectively.

Next note that the computation of $\pbest$ and $\nbest$ is not seen in the
update equations~\eqref{eq:velupdate} and \eqref{eq:posupdate}, so we will make
this mathematically explicit.  This update is usually implemented as a pair of
independent ``if``-statements, but that is a bit awkward for our purposes
so we instead use a set of indicator variables, one each representing each
possible update to the \emph{pair} $\casedef$.  These indicator variables allow
us to easily identify the cases in which updates will be needed for $\nbest$,
$\pbest$ or both.

For example, we can write $\casexn$ to identify the case where $\pbest$ will be
updated with the better position $\ppos_{i,t}$, shown in the first positions,
but a no new neighborhood best was found so $\nbest$ remains in the second
position.

For any topology with 2 neighbors there are 7 possible update cases listed in
Table~\ref{tab:evals}.  This list assumes, as is typically the case, that each
particle is also part of its own neighborhood (can cause an update to
$\nbest$). The list also omits the case where $\nbest$ is updated by
$\ppos_{t}$ but $\pbest$ was not updated, as this is logically impossible.
$\caseset$ represents the set of all update cases.

For minimization, we can define this example indicator variable representing the case
where the particle will need to update $\pbest$ with the better
position $x_{t}$ for the next iteration ($t+1$), but $\nbest$ will not updated by:
\begin{align}
\nonumber
	\indic_{t+1}^{\casexn} & (\ofunc ( \ppos_{t+1} ) ,\ofunc(\leftn_{t+1}),\ofunc(\rightn_{t+1}) ,\ofunc(\pbest_{t}) ,\ofunc(\nbest_{t}))= \\
\nonumber
	& 1 \ if \  \ofunc(\ppos_{t+1}) > \ofunc(\pbest_{t}) \\
\nonumber
	& \quad \quad \ \wedge \ \ofunc(\nbest_{t}) > \ofunc(\ppos_{t+1}) \\
\nonumber
	& \quad \quad \ \wedge \ \ofunc(\nbest_{t}) > \ofunc(\leftn_{t+1}) \\
\nonumber
	& \quad \quad \ \wedge \ \ofunc(\nbest_{t}) > \ofunc(\rightn_{t+1}) \\
\label{eq:deficasexn}
	& 0 \ otherwise
\end{align}

@@

For each cases identified in Table~\ref{tab:evals} and identified by a
corresponding indicator variable defined above, we also provide notation for the \emph{next} velocity
update equation; once again using $\casexn$ as an example:
\begin{align}
\nonumber
	\specvel_{t+1}^{\casexn} & (\pvel_t, \ppos_{t+1}, \leftn_{t+1}, \rightn_{t+1}, \pbest_{t},
        \nbest_{t}, \pURand_{t}, \nURand_{t}) \\
\label{eq:defvocasexn}
		&= \constriction \bigl[ \pvel_{t} +
			\pcoeff\pURand_{t+1}\otimes(\pbest_{t+1} - \ppos_{t+1}) \\
		& \quad \quad \quad \; + \ncoeff\nURand_{t+1}\otimes(\nbest_{t+1} - \ppos_{t+1})
		\bigr]
\end{align}

Again we will drop the parameters for compactness and since
they can be inferred from the subscripts on $\specvel_{t}^{\casexn}$.
The update equation is taken exactly from the standard update equation~\eqref{eq:velupdate}, but
since it is restricted to a specific case, $\casexn$ for this example,
it is no longer necessary to use $\pbest$ and $\nbest$, since in this case
$\pbest_{t+1}=\ppos_{t+1}$ and $\nbest_{t+1}=\nbest_{t}$ we can re-write
$\specvel_{t+1}^{\casexn}$ as:
\begin{align}
\nonumber
	\specvel_{t+1}^{\casexn} &= \constriction \bigl[ \pvel_{t} +
			\pcoeff\pURand_{t+1}\otimes(\ppos_{t+1} - \ppos_{t+1}) \\
\label{eq:defvcasexn}
			& \quad \quad \quad \; + \ncoeff\nURand_{t+1}\otimes(\nbest_{t} - \ppos_{t+1})
		\bigr]
\end{align}

For each cases identified in
Table~\ref{tab:evals}
and identified by a corresponding indicator variable defined above,
we define also a position update equation with its compact form; once again using $\casexn$ as an example:
\begin{align}
\label{eq:defpcasexn}
	\specpos_{t+1}^{\casexn} & (\ppos_{t}, \pvel_{t},\ppos_{t+1} ,\leftn_{t+1},\rightn_{t+1} ,\pbest_{t} ,\nbest_{t})= \\
\nonumber
	& = \specpos_{t+1}^{\casexn} =  \ppos_{t} + \specvel_{t+1}^{\casexn}
\end{align}
We call these values, $\specpos^{c}) \ c \in\ \caseset$ (7 of them in this case where each particle has 2 neighbors) as ``speculative children''
since they will only be useful if the corresponding $\indic^{c} \ c \in \caseset$ has the value 1.

@@

Returning our attention to the computtaion of $\specpos$ in \eqref{eq:pos2update}and writing it with the parameters which were omitted above,
we obtain:
\begin{align}
\nonumber
\ppos_{t+1} &= \sum_{c \in \caseset} \\
\nonumber
& \specpos_{t+1}^{c}(\ppos_{t+1},\leftn_{t+1},\rightn_{t+1},
\pbest_{t},\nbest_{t},\pURand_{t}, \nURand_{t}) \\
\label{eq:val2updatelong}
& \indic_{t}^{c}(\ofunc ( \ppos_{t+1} ) ,\ofunc(\leftn_{t+1}),\ofunc(\rightn_{t+1}) ,\ofunc(\pbest_{t}) ,\ofunc(\nbest_{t}))
\end{align}
In this form the important point to notice is that none of the $\specpos_{t+1}^{c} \ c \in\ \caseset$ depend upon $f(x_t)$
or any other objective function evaluation at $t$.
Thus PSO has been re-factored such that the algorithm can begin computing
all 7 of the objective function evaluations potentially needed in iteration $t+1$ before $f(x_t)$ is computed.

This refactoring comes a significant expense, instead of just computing
$\ofunc(\ppos_{i,t})$ and then $\ofunc(\ppos_{i,t+1})$ in the usual way, this refactored PSO
computes $\ofunc(\ppos_{t})$ and all 
of the $\ofunc(\specpos_{t+1}^{c}) \ \ c \in \caseset$ at the same time.
Once the evaluation of $f(\ppos_{t})$ is completed only one
of the indicator functions $\indic_{t}^{c} \ c \in \caseset$ will be set to one.
We will refer to the indicator function which is set to one as the
\emph{behavior preserving indicator} or $bp$, since it indicates which of
the computations $\ofunc(\specpos_{t}^{c}) \  c \in \caseset$ preservers the behavior of PSO.
The rest 
computations $\ofunc(\specpos_{t}^{c}) \ c \in \caseset-\{bp\}$
will be ignored, and might just as well never have been computed.
We call these evaluations of $\ofunc(\specpos_{t+1}^{c})$ ``speculative children''
because they may or may not be needed.

The potential value of these speculative children lies in the phrase used above
``this re-factored PSO
can compute $\ofunc(\ppos_{t+1})$ and all 
of the $\ofunc(\specpos_{t+1}^{c}) \ \ c \in \caseset$ at the same time''. 
Remember that our intent is to accelerate parallel PSO and that we
are interested in objective functions that have at least non-trivial computational
complexity.
That is, we assume that computing
$\ofunc(\ppos_{t})$ and all 
of the $\ofunc(\specpos_{t+1}^{c}) \ c \in \caseset$ are the computationally difficult parts.

For example, suppose that unlimited processors were available and that the evaluation of 
the objective function took 1 hour. Suppose also that for each particle
the computations
$\ofunc(\ppos_{t})$ and all 
of the $\ofunc(\specpos_{t+1}^{c}) \ c \in \caseset$ were sent to separate
processors and completed in parallel.
\emph{Given} the values 
of $\ofunc(\ppos_{t})$ for all particles
the computation of the indicator functions $\indic_{t}^{c} \ c \in \caseset$ and to complete iteration t+1
is inconsequential relative to the computational cost of obtaining the function values. 
In this environment all of the function evaluations needed for 2 iterations of PSO
could be completed in just one hour where as the conventional implementation would
require two hours for those same 2 iterations of PSO.

>>>>

Particle swarm optimization (PSO) has previously been parallelized primarily by 
distributing the computation corresponding to particles or sets
across multiple processors.  However, in many cases this approach is not efficient when scale to 
very large populations sizes (on very large clusters).
In this paper we present a different way of using multiple processors of the 
computation in PSO, motivated by the ``speculative execution'' in modern computer hardware and which we call
Speculative Evaluation in PSO (SEPSO).

In SEPSO future positions of the particles are
speculated and evaluated in parallel with the evaluation of current positions.
We show that a set of separate speculative evaluation threads can be constructed
such that the set will always include the position and corresponding
evaluation that would normally have occurred at the next iteration of standard PSO.
Thus SEPSO performs two
iterations of PSO at once, allowing it to achieve better fitness faster.
Not only can SEPSO can achieve better fitness in the same number of iterations,
it also reaches better fitness for the same number of total function evaluations
in large scale parallel environments.

Speculative Evaluation in PSO can be enhanced by diverging from the
behavior of the standard PSO to make better use of the information
obtained by the speculative evaluations.
Using these algorithmic modifications, we see dramatic improvements in the
effectiveness of the algorithm. 
\end{abstract}

@@

These previous parallel techniques 
distribute the computation needed by the particles in the swarm over the available processors.
if more processors are available, these prior techniques increase the
number of particles in the swarm, either by adding individual particles or by
adding entire new sub-swarms but in all cases
the number of processors never exceeds the number of particles.
The number of iterations of PSO that the
algorithms can perform is thus inherently limited by the time it takes to
evaluate the objective function---additional processors add more particles, but do not
make the iterations go any faster.

The purpose of this paper is to consider PSO parallelization strategies for
clusters of thousands of processors and functions for which a single
evaluation will take at least tens of seconds, but probably hours or perhaps days.
Our purpose is to explore the question ``Since PSO is often run with
a swarm of size of 50-100, what would one do if a few thousand processors were available?''

For many functions there comes a point of diminishing returns with respect to adding particles.
In \fig{basic-sphere} we show the value obtained after 20,000 function evaluations (not iterations)
as a function of swarm size
for the well-know benchmark function Sphere (20 dimensions, reporting the average of thirty runs). 
Increasing the swarm size from 10 to 50 has a significant effect value obtained
however, increasing the swarm size from 100 to 200 makes the algorithm less efficient,
that is, it reduces the progress the algorithm makes per-evaluation.
For this reason, previous work has recommended the use a swarm size of 50 for this function
~\cite{bratton-sis07}.
Thus, in at least some cases, added particles indefinitely will not yield an efficient implementation.

@@

To solve this problem, we take inspiration from speculative execution
techniques commonly used in processors.  Modern processors, when faced with a
branch on which they must wait (e.g. a memory cache miss), guess which way the
branch will go and start executing, being careful not to make changes that can not be undone.
If the processor guesses right, execution
is much farther ahead than if it had just idly waited on the memory reference.  If it
guesses wrong, execution restarts where it would have been anyway.
This is called speculative execution or branch prediction.

\begin{quote}
SKIP SKIP

No really, try reading without this! The idea of speculative execution can be applied to some types of optimization
algorithms.  When the next sampling position of the algorithm does not depend
on the value of the objective function at the current position,
future positions can be speculated and evaluated in parallel with the current
position.
This is trivially true for optimization by random sampling, for example.
The next random sample does not depend on the previous point nor on the value
of the objective function at that point in any way.
It depends only on a random number.
There is no need to wait to see
what objective function value was obtained at the previous point,
evaluation of the object function at the next point can begin as soon as the next random number is drawn.
In other optimization algorithms, the next position depends on the previous position, but not the
value obtained there, a random walk for example.
In this case
the next position to be evaluated
depends only on the previous position and a random number.
In this case, as it was in the case of random sampling,
the object function evaluation of the next position can be done in parallel
because it 
does not depend on the \emph{value} of the objective function at the previous position.
There are an infinite number of possible next positions, but only one given the draw of the random number.

In still other cases, the next position to be evaluated may depend on the current position and some
discrete valued (binary, for example) test on the value obtained at the current position.
For example, consider a greedy random walk, in which the state only advances if a better
position is found.
That is, if the current value is better than the previous value,
use the sum of the current position and the random number as the next position.
Otherwise, computer the next position using the sum of the previous position and the random number.
In this case the objective function can be evaluated at three positions:
(1) the value of the objective function at the current position,
(2) the value at the position if the test turns out to be true and (3) the value at the position
if the test turns out to be false; all in parallel.
In this example it is important to note that we do not need the value of the objective function at the current
point to to reduce the infinite number of possible next positions to just two alternatives.
Choosing between these two alternatives can be done post-hoc, after the objective function value
at the current position and at the two alternative are computed.
For all but the most computationally-trivial objective functions, this approach saves almost half of the
``wall-clock'' time.
This approach has been applied to simulated annealing~\cite{witte-tpds91} which like the greedy random walk,
depends on positions and a binary test.

This approach can not be used in gradient methods where the
step size for sampling typically depends on the gradient and therefor the position and 
a set of infinitely many possible \emph{values} outcomes
obtained at the current position (not binary or even finite).
\end{quote}

@@

In this paper we show that the results of standard PSO can be reproduced \emph{exactly}, two iterations at a time, using
a speculative approach like the ones described above. We prove that the standard PSO equation can
be factored such that a small set of speculative positions can be found which will \emph{always}
include the position computed in the next iteration.
By computing the value of the objective function for each of the speculative positions at
the same time the algorithm evaluates the objective function for the current position,
It is possible to know the objective function values for both the current and the next iteration at the same time.
We demonstrate this principle by implementation and show that it
produces exactly the same results as standard PSO, but two iterations at a time.
The resulting implementation runs efficiently on large clusters where the number of processors is much larger than
a typical or reasonable number of particles. Producing better results in less ``wall-clock'' time and,
surprisingly, with fewer total function evaluations.

Furthermore, we show that if we relax the requirements of the algorithm, no
longer demanding that it strictly reproduce the exact behavior of standard PSO, we
can introduce new speculative techniques that often out-perform PSO.
These variants make better use of the extra information obtained from
the extra exploration made by the speculative function evaluations.
We also explore the idea that, like branch prediction in processors, we
need not speculatively evaluate \emph{all} possible future positions,
we can accelerate the algorithm even if we are just \emph{likely} to have guessed right.
By pruning the speculation to just paths that are statistically likely to reproduce
the paths that are equivalent to PSO we can also 
speculating several iterations ahead instead of just one.
We also consider several recovery strategies for cases where
the pruned set of speculative evaluations does not contain the evaluation that standard PSO would have done.

@@

At each iteration ($t+1$) of the algorithm, the position $\ppos_t$ and velocity
$\pvel_t$ of each particle are updated as follows:
\begin{align}
\nonumber
	\pvel_{t+1} &=
		\constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\label{eq:velupdate}
			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
		\bigr] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1}
\end{align}
where \( \nURand_{t} \) and \( \nURand_{t} \) are vectors of independent
and identically distributed random numbers drawn from a uniform
distribution, the \( \otimes \) operator is an element-wise vector
multiplication, $\pbest$ (called personal best) is the best position the
current particle has seen, and $\nbest$ (called neighborhood best) is the best
position the neighbors of the current particle have seen~\cite{bratton-sis07}.  \( \ncoeff \), \(
\pcoeff \), and \( \constriction \) are parameters with prescribed values
required to ensure convergence (2.05, 2.05, and .73,
respectively)~\cite{clerc-tec02}. Note also that we use a subscript for the 
time step $t$ or $t+1$ and have moved all other qualifiers,
such as the personal qualifier ``$\pers$'' to the superscript position.
Superscripts can be miss read as exponents, but they are qualifiers throughout this paper.
We use $p$ to indicate the number of particles in the swarm.

There are many ways of defining the neighbors of a particle. 
Changing the
neighborhood topology has a significant effect on the performance of the algorithm and 
this effect varies widely by function type.
%, as would be expected by the No Free Lunch
%theorem~\cite{wolpert-tec97}.
The two most common topologies used in the
literature are the Ring topology and the Complete
topology~\cite{bratton-sis07}.  In the Ring topology, each particle has one
(or more) neighbors to either side of it; in the Complete topology, every particle is a
neighbor to every other particle.  It is generally assumed that, a particle is also a
neighbor to itself in that its own position and value is considered when updating the
particle's personal best, $\nbest$.  Thus with $p$ particles, using the Ring topology each
particle has three neighbors: $i-1$, $i$ (itself), and $i+1$.  With the Complete
topology, each particle has $p$ neighbors.
% TODO USE THIS LATER!
% In this paper we use these
% topologies, as well as a parallel adaptation of the Complete topology, called
% Random, that has been shown to simulate the behavior of Complete with far less
% communication~\cite{mcnabb-cec09}.  In the Random topology, each particle
% randomly picks two other particles to share information with at each iteration,
% along with itself.  Thus in both the Ring and the Random topologies, all
% particles have 3 neighbors.

@@

To address this issue we present here a different use of additional processors.
We take inspiration from speculative execution in processors and instead of adding particles
we will employ a speculative evaluation approach like those described in the introduction. 
This allows us to perform iterations two at a time.

To do so we must first re-factor PSO such that the determination of the $\pbest$ and $\nbest$,
which depend upon the value of the objective function, is sepearate from the rest of the computation.
%First consider formula for the \emph{next} iteration of PSO, which is identical
%to Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate} except the time
%subscripts are incremented.  As noted above and
For simplicity in this
explanation, we restrict our discussion here to the case of a Ring topology
with two neighbors which we will call the ``right neighbor'' and ``left
neighbor'' neighbors identified for each particle as $\rightn$ and $\leftn$
respectively.

Next note that the computation of $\pbest$ and $\nbest$ is not seen in the
update equations~\eqref{eq:velupdate} and \eqref{eq:posupdate}, so we will make
this mathematically explicit.  This update is usually implemented as a pair of
independent ``if``-statements, but that is a bit awkward for our purposes
so we instead use a set of indicator variables, one each representing each
possible update to the pair $\casedef$.  These indicator variables allow
us to easily identify the cases in which updates will be needed for $\nbest$,
$\pbest$ or both.

For a topology with 2 neighbors there are 7 possible update cases identified in
Table~\ref{tab:evals}.
For example and as indicated in the table, we write $\casexn$ to represent the case where $\pbest$ will be
updated with the better position $\ppos_{i,t}$, shown in the first positions,
but a no new neighborhood best was found so $\nbest$ remains in the second
position.
%This list assumes, as is typically the case for the ring topology, that each
%particle is also part of its own neighborhood (can cause an update to
%$\nbest$).
The list omits the case where $\nbest$ is updated by
$\ppos_{t}$ but $\pbest$ was not updated, as this is logically impossible.
$\caseset$ represents the set of all possible update cases.

For minimization, we can define the indicator variable representing the case $\casexn$,
where the particle will need to update $\pbest$ with the better
position $x_{t}$, but $\nbest$ will not updated, by:
\begin{align}
\nonumber
	\indic_{t+1}^{\casexn} & (\ofunc ( \ppos_{t+1} ) ,\ofunc(\leftn_{t+1}),\ofunc(\rightn_{t+1}) ,\ofunc(\pbest_{t}) ,\ofunc(\nbest_{t}))= \\
\nonumber
	& 1 \ if \  \ofunc(\ppos_{t+1}) > \ofunc(\pbest_{t}) \\
\nonumber
	& \quad \quad \ \wedge \ \ofunc(\nbest_{t}) > \ofunc(\ppos_{t+1}) \\
\nonumber
	& \quad \quad \ \wedge \ \ofunc(\nbest_{t}) > \ofunc(\leftn_{t+1}) \\
\nonumber
	& \quad \quad \ \wedge \ \ofunc(\nbest_{t}) > \ofunc(\rightn_{t+1}) \\
\label{eq:deficasexn}
	& 0 \ otherwise
\end{align}

@@

For each cases identified in Table~\ref{tab:evals} and identified by a
corresponding indicator variable defined above, we also provide notation for the velocity
update equation; once again using $\casexn$ as an example:
\begin{align}
\nonumber
	\specvel_{t+1}^{\casexn} & (\pvel_t, \ppos_{t+1}, \leftn_{t+1}, \rightn_{t+1}, \pbest_{t},
        \nbest_{t}, \pURand_{t}, \nURand_{t}) \\
\label{eq:defvocasexn}
		&= \constriction \bigl[ \pvel_{t} +
			\pcoeff\pURand_{t+1}\otimes(\pbest_{t+1} - \ppos_{t+1}) \\
		& \quad \quad \quad \; + \ncoeff\nURand_{t+1}\otimes(\nbest_{t+1} - \ppos_{t+1})
		\bigr]
\end{align}

Again we will drop the parameters for compactness and since
they can be inferred from the subscripts on $\specvel_{t}^{\casexn}$.
The update equation is taken exactly from the standard update equation~\eqref{eq:velupdate}.
However, since $\specvel_{t}^{\casexn}$ is restricted to the specific case, $\casexn$,
it is no longer necessary to use $\pbest$ and $\nbest$, since when restricted to this case
$\pbest_{t+1}=\ppos_{t+1}$ and $\nbest_{t+1}=\nbest_{t}$ and we can re-write
$\specvel_{t+1}^{\casexn}$ as:
\begin{align}
\nonumber
	\specvel_{t+1}^{\casexn} &= \constriction \bigl[ \pvel_{t} +
			\pcoeff\pURand_{t+1}\otimes(\ppos_{t+1} - \ppos_{t+1}) \\
\label{eq:defvcasexn}
			& \quad \quad \quad \; + \ncoeff\nURand_{t+1}\otimes(\nbest_{t} - \ppos_{t+1})
		\bigr]
\end{align}

In the same way we can create notation for the position update
equation~\eqref{eq:velupdate}; once again using $\casexn$ as an example:
\begin{align}
\label{eq:defpcasexn}
	\specpos_{t+1}^{\casexn} & (\ppos_{t}, \pvel_{t},\ppos_{t+1} ,\leftn_{t+1},\rightn_{t+1} ,\pbest_{t} ,\nbest_{t})= \\
\nonumber
	& = \specpos_{t+1}^{\casexn} =  \ppos_{t} + \specvel_{t+1}^{\casexn}
\end{align}
We call these values, $\specpos^{c}) \ c \in\ \caseset$ (7 of them in this case where each particle has 2 neighbors plus itself)
as ``speculative children''
since they will only be useful if the corresponding $\indic^{c} \ c \in \caseset$ has the value 1.

@@

Returning our attention to the computation of $\specpos$ in \eqref{eq:pos2update} and writing it
with the parameters which were omitted above,
we obtain:
\begin{align}
\nonumber
\ppos_{t+1} &= \sum_{c \in \caseset} \\
\nonumber
& \specpos_{t+1}^{c}(\ppos_{t+1},\leftn_{t+1},\rightn_{t+1},
\pbest_{t},\nbest_{t},\pURand_{t}, \nURand_{t}) \\
\label{eq:val2updatelong}
& \indic_{t}^{c}(\ofunc ( \ppos_{t+1} ) ,\ofunc(\leftn_{t+1}),\ofunc(\rightn_{t+1}) ,\ofunc(\pbest_{t}) ,\ofunc(\nbest_{t}))
\end{align}
In this form the important point to notice is that there are only 7 values (for this ring topology) in the set formed by
$\specpos_{t+1}^{c} \ c \in\ \caseset$ and that none of them depend upon $f(x_t)$
or any other objective function evaluation at $t$.
Thus PSO has been re-factored such that the algorithm can begin computing
all 7 of the objective function evaluations potentially needed in iteration $t+1$ \emph{before} $f(x_t)$ is computed.
Once the evaluation of $f(\ppos_{t})$ is completed for all particles only one
of the indicator functions $\indic_{t}^{c} \ c \in \caseset$ will be set to one.
We will refer to the indicator function which is set to one as the
``behavior preserving indicator'' or $bp$, since it indicates which of
the computations $\ofunc(\specpos_{t}^{c}) \  c \in \caseset$ preservers the behavior of PSO.
The rest 
computations $\ofunc(\specpos_{t}^{c}) \ c \in \caseset-\{bp\}$
will be ignored, and might just as well never have been computed.
We call these evaluations of $\ofunc(\specpos_{t+1}^{c})$ ``speculative children''
because they may or may not be needed.

This refactoring comes a significant expense, instead of just computing
$\ofunc(\ppos_{i,t})$ and then $\ofunc(\ppos_{i,t+1})$ in the usual way, this refactored PSO
computes $\ofunc(\ppos_{t})$ and all 
of the $\ofunc(\specpos_{t+1}^{c}) \ \ c \in \caseset$, 8 function evaluations in this case.
The potential value of these speculative children lies in the phrase used above
``this re-factored PSO
can compute $\ofunc(\ppos_{t+1})$ and all 
of the $\ofunc(\specpos_{t+1}^{c}) \ \ c \in \caseset$ at the same time''. 
Remember that our intent is to accelerate parallel PSO and that we
are interested in objective functions that have at least non-trivial computational
complexity.
That is, we assume that computing
$\ofunc(\ppos_{t})$ is the computationally difficult part of one iteration of PSO.

The value of the refactoring can best be seen in a small example.
Suppose that unlimited processors were available and that the evaluation of 
the objective function took 1 hour. Suppose also that for each particle
the computations
$\ofunc(\ppos_{t})$ and all 
of the $\ofunc(\specpos_{t+1}^{c}) \ c \in \caseset$ were sent to separate
processors and completed in parallel.
\emph{Given} the values 
of $\ofunc(\ppos_{t})$ for all particles
the computation of the indicator functions $\indic_{t}^{c} \ c \in \caseset$ and to complete iteration t+1
is inconsequential relative to the computational cost of obtaining the function values. 
In this environment all of the function evaluations needed for 2 iterations of PSO
could be completed in just one hour where as the conventional implementation would
require two hours for those same 2 iterations of PSO.

While the algebra shows that the computation of 
equations~\eqref{eq:vel2update} and \eqref{eq:pos2update} are identical to the 
equations~\eqref{eq:velupdate} and \eqref{eq:posupdate},
There are some issues better seen in an algorithmic setting, in particular the use of the random
vectors, 
$\pURand_{t}$ and $\nURand_{t}$.
As with the algebra given above, each transformation (used below) of the algorithm is
provably equivalent; except that the algorithms will no longer be able to do an odd number of iterations,
since it will do two at a time. In the algorithmic transformations given below,
instructions are moved around and added, but the
resultant algorithm, produces numerically identical results.
